---
title: "Final Project: Public Safety Measures and Public Health for Covid-19"
author: "Paul Croome, Rodrigo Rosales Alvarez, Ann Siddiqui and Kane Smith"
date: "2022-12-14"
output: pdf_document
subtitle: "Data 603 - Statistical Modelling with Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{center}\textbf{} 
\\Professor: Dr. Paul Galpern
\\University of Calgary
\\Calgary, Alberta
\end{center}



\newpage
\tableofcontents



\newpage
# Introduction
The domain of our project covers healthcare-related indicators of the wellbeing of countries during the coronavirus disease 2019 (COVID-19) pandemic. In particular, we will be examining data related to the prevalence and severity of the COVID-19 pandemic and the governmental and societal measures taken to reduce the spread of the disease. These data were all daily reported between January 2020 and October 2022.

This is an interesting and important topic of study because, in our increasingly interconnected world, contagious diseases can be transmitted over vast distances remarkably easily. Even small, remote outbreaks of diseases anywhere in the world can swiftly turn into a global pandemic, which can then cause devastation on personal, societal, and worldwide scales. 


## Research Questions
1) What population-related metrics of countries around the world are most strongly related to the prevalence and severity of COVID-19 experienced in a country between February 2020 and October 2022 (as measured by average daily COVID-19 cases and deaths)? 

    a) What is the best model that can be built from these data for predicting the average daily new COVID-19 cases experienced in a country? 
    
    b) What is the best model that can be built from these data for predicting the average daily new COVID-19 deaths experienced in a country?

2) Among countries with reliably reported data relating to cases, positive test rates, vaccinations, and boosters, what societal and governmental responses to the COVID-19 pandemic are most strongly related to the prevalence and severity of COVID-19 experienced in a country between February 2020 and October 2022 (as measured by average daily COVID-19 cases and deaths)?

    a) What is the best model that can be built from these daily-reported data for predicting the daily new COVID-19 cases in a country?
    
    b) What is the best model that can be built from these daily-reported data for predicting the daily new COVID-19 deaths in a country?



\newpage
# Data Set Definition
The dataset we will use consists of diverse information related to the COVID-19 pandemic, including a country’s daily rates of COVID-19 diagnoses, hospitalizations, deaths, vaccinations, and booster shots. We will use features of these data to determine the prevalence and severity of the COVID-19 pandemic for each country . The dataset consists of daily information from January 1, 2020 to October 26, 2022 for more than 220 countries; each row corresponds to the Covid-19 information reported by an specific country in a certain date.

This dataset is contained in a CSV file and is licensed for open access under the Creative Commons BY license. The data set was put together by [Our World in Data](https://ourworldindata.org/coronavirus); more important the data set is being update daily by the same organization, for more information about the data pipeline and how the data set is being maintained click [here](https://docs.owid.io/projects/covid/en/latest/data-pipeline.html#overview).  



\newpage
# Methodology

## Libraries 
```{r, results='hide', echo=FALSE, include=FALSE}
library(dplyr) # data manipulation
library(ggplot2) # graphs
library(GGally) # extension to ggplot2
library(mosaic) # statistics and calculus functions
library(tidyverse) # data science 
library(tidyr) # tidy data
library(olsrr) # teaching and learning OLS regression
library(leaps) # best subset of variables for a model
library(mctest) # VIF
library(car) # VIF
library(lmtest) # Bracuh-Pagan
library(agricolae) # Newman-Keuls
library(MASS) # box-cox
options(scipen=999) # show results withour cientific notation
```

## Data Import
As our data is in CSV format, we simple use the function **read_csv()** to import our file into a data frame.
```{r}
covid_raw <- read_csv('data.csv', show_col_types = FALSE)
head(covid_raw, 5)
```

## Data Cleaning 
Many countries and facilities are under reporting Covid-19 statistics like cases and deaths, according to Claire Klobucista from Council of Foreign Relations. This is could be catastrophically as government could not respond accordingly to the real situation. For our research this is very important as well, if we put bad data into our model, we would create a bad model. To solve this problem we will remove from our dataset the bottom 5% of the countries for number of new cases smoothed per million and number of new deaths smoothed per million. 
```{r}
# Getting countries with bottom 5% of new_cases_smoothed_per_million
tenth_percentile_cases <- quantile(covid_raw$new_cases_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_cases <- covid_raw[covid_raw$new_cases_smoothed_per_million < tenth_percentile_cases,]
bad_country_list_cases <- unique(bad_country_cases$location)

# Getting countries with bottom 5% of new_deaths_smoothed_per_million
tenth_percentile_deaths <- quantile(covid_raw$new_deaths_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_deaths <- covid_raw[covid_raw$new_deaths_smoothed_per_million < tenth_percentile_deaths,]
bad_country_list_deaths <- unique(bad_country_deaths$location)

# Remove countries that appear in either above lists
bad_country_list <- append(bad_country_list_deaths, bad_country_list_cases)
good_countries <-   covid_raw[!covid_raw$location %in% bad_country_list, ]
```

After doing this we realized




## Variable Definition.

*Independent Variable*
1) new_cases: new confirmed cases of COVID-19. Continuous Variable.
2) new_deaths: new deaths attributed to COVID-19. Continuous Variable.

*Dependent Variables*


\newpage
# Results



\newpage
# Discussion



\newpage
# References
Klobucista, Claire (2021, May 10). By How Much Are Countries Underreporting COVID-19 Cases and Deaths?. Council of Foreign Relations. https://www.cfr.org/in-brief/how-much-are-countries-underreporting-covid-19-cases-and-deaths


