---
title: "P603_Rodrigo"
author: "Rodrigo Rosales"
date: "2022-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide', echo=FALSE, include=FALSE}
# Import libraries
library(dplyr)
library(ggplot2)
library(mosaic)
library(tidyverse)
library(resampledata)
library(tidyr)
library(olsrr)
library(leaps)
library(GGally)
options(scipen=999)
```

```{r}
# Importing COVDI-19 data set
covid_raw <- read_csv('data.csv', show_col_types = FALSE)
```

```{r}
# Getting countries with bottom 5% of new_cases_smoothed_per_million
tenth_percentile_cases <- quantile(covid_raw$new_cases_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_cases <- covid_raw[covid_raw$new_cases_smoothed_per_million < tenth_percentile_cases,]
bad_country_list_cases <- unique(bad_country_cases$location)

# Getting countries with bottom 5% of new_deaths_smoothed_per_million
tenth_percentile_deaths <- quantile(covid_raw$new_deaths_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_deaths <- covid_raw[covid_raw$new_deaths_smoothed_per_million < tenth_percentile_deaths,]
bad_country_list_deaths <- unique(bad_country_deaths$location)

# Remove countries that appear in either above lists
bad_country_list <- append(bad_country_list_deaths, bad_country_list_cases)
good_countries <-   covid_raw[!covid_raw$location %in% bad_country_list, ]
```

```{r}
# change here to use final cleaned dataset
covid_data <- good_countries
covid_data$smokers <- (covid_data[['male_smokers']] + covid_data[['female_smokers']]) / 2

# dropping columns
covid = subset(covid_data, select = -c(iso_code, date, continent, handwashing_facilities, tests_units) )

# NaN to 0
covid[is.na(covid)] = 0

# Aggregation
covid_agg <- covid %>% group_by(location) %>% summarise(across(everything(), mean), .groups = 'drop') %>% as.data.frame()
```


# CASES MODEL
```{r}
model_cases_full = lm(new_cases ~ aged_65_older + smokers + cardiovasc_death_rate + diabetes_prevalence + extreme_poverty + gdp_per_capita + median_age + life_expectancy + population + stringency_index + human_development_index + reproduction_rate + hosp_patients + new_tests + population_density, data=covid_agg)

summary(model_cases_full)
```
```{r}
model_cases_stepwise = ols_step_both_p(model_cases_full, pent=0.1, prem=0.3, progress=TRUE)
```
```{r}
model_cases_revised = lm(new_cases ~ population + hosp_patients + life_expectancy, data=covid_agg)

summary(model_cases_revised)
```

```{r}
model_cases_interactions = lm(new_cases ~ (population + hosp_patients + life_expectancy)^2, data=covid_agg)
summary(model_cases_interactions)
```

```{r}
model_cases_interactions_2 = lm(new_cases ~ population + hosp_patients + life_expectancy + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_interactions_2)
```

```{r}
new_data <-data.frame(covid_agg$new_cases, covid_agg$population, covid_agg$hosp_patients, covid_agg$life_expectancy)

ggpairs(new_data, lower = list(continuous="smooth_loess", combo="facethist", discrete="facetbar", na="na"))
```

```{r}
model_cases_squared = lm(new_cases ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_squared)
```

```{r}
model_cases_final = lm(new_cases ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_final)
```



# MODEL DEATHS
```{r}
model_deaths_full = lm(new_deaths ~ aged_65_older + smokers + cardiovasc_death_rate + diabetes_prevalence + extreme_poverty + gdp_per_capita + median_age + life_expectancy + population + stringency_index + human_development_index + reproduction_rate + hosp_patients + new_tests + population_density, data=covid_agg)

summary(model_deaths_full)
```

```{r}
model_deaths_stepwise = ols_step_both_p(model_deaths_full, pent=0.1, prem=0.3, progress=TRUE)
```

```{r}
model_deaths_revised = lm(new_deaths ~ population + hosp_patients + life_expectancy, data=covid_agg)

summary(model_deaths_revised)
```

```{r}
model_deaths_interactions = lm(new_deaths ~ (population + hosp_patients + life_expectancy)^2, data=covid_agg)
summary(model_deaths_interactions)
```


```{r}
model_deaths_interactions_2 = lm(new_deaths ~ population + hosp_patients + life_expectancy + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_interactions_2)
```

```{r}
new_data <-data.frame(covid_agg$new_deaths, covid_agg$population, covid_agg$hosp_patients, covid_agg$life_expectancy)

ggpairs(new_data, lower = list(continuous="smooth_loess", combo="facethist", discrete="facetbar", na="na"))
```

```{r}
model_deaths_squared = lm(new_deaths ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_squared)
```

```{r}
model_deaths_final = lm(new_deaths ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_final)
```

# ASSUMPTIONS 

## CASES MODEL

### NORMALITY 

Testing for normality using the Shapiro-Wilk test:

**Null hypothesis:** the sample data are significantly normally distributed

**Alternative hypothesis:** the sample data are not significantly normally distributed

We will set the alpha value to 0.05.

```{r}
shapiro.test(residuals(model_cases_final))
```


Plotting a Q-Q plot:

```{r}
ggplot(q3_final, aes(sample=model_cases_final$residuals)) +
stat_qq() +
stat_qq_line()
```

From the output of our test, we get a p-value of 0.00000000000000022 which is less than 0.05. This means we can reject our null hypothesis that our sample data is significantly normally distributed and conclude with a significance level of 0.05 that our sample data is not normally distributed. This means that there is a problem with the normality assumption.

### HOMOSCEDASTICITY

Testing for heteroscedasticity using the Breusch-Pagan test:

**Null hypothesis:** heteroscedasticity is not present ($H_0: \sigma^2_1 = \sigma^2_2 =... = \sigma^2_n$)

**Alternative hypothesis:** heteroscedasticity is present ($H_a:$ at least one $\sigma^2_i$ is different from the others)

We will set the alpha value to 0.05.

```{r}
bptest(model_cases_final)
```

From the output of our test, we get a p-value of 0.0000001319 which is less than 0.05. This means we can reject the null hypothesis that there is homoscedasticity and conclude with a significance level of 0.05 that our model is not homoscedastic. This means that there is a problem with the homoscedasticity assumption. 

### LINEARITY 

Testing for linearity using a residuals vs predicted $\hat{Y}$ plot:
```{r}
ggplot(model_cases_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, our model does seem to be linear. ??????

### OUTLIERS

Testing for outliers using leverage:
```{r}
lev=hatvalues(model_cases_final)
p = length(coef(model_cases_final))
n = nrow(covid_agg)
outlier3p = lev[lev>(3*p/n)]
outlier3p
```

```{r}
plot(model_cases_final,which=5)
```

From our output, there seems to be two data points, row 12, 44, 64, 76, 89, 99, 101, 107, and 110 with a leverage greater than $3p/n$. 

### INDEPENDANT ERRORS

Plotting residuals vs. fitted values to see if the error terms look independant from each other:
```{r}
ggplot(model_cases_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```


Looking at the plot above, it seems that the residuals are evenly distributed, meaning that the residuals are independent from each other. ??????

## DEATHS MODEL

### NORMALITY 

Testing for normality using the Shapiro-Wilk test:

**Null hypothesis:** the sample data are significantly normally distributed

**Alternative hypothesis:** the sample data are not significantly normally distributed

We will set the alpha value to 0.05.

```{r}
shapiro.test(residuals(model_deaths_final))
```

Plotting a Q-Q plot:

```{r}
ggplot(q3_final, aes(sample=model_deaths_final$residuals)) +
stat_qq() +
stat_qq_line()
```

From the output of our test, we get a p-value of 0.00000000000000022 which is less than 0.05. This means we can reject our null hypothesis that our sample data is significantly normally distributed and conclude with a significance level of 0.05 that our sample data is not normally distributed. This means that there is a problem with the normality assumption.

### HOMOSCEDASTICITY 

Testing for heteroscedasticity using the Breusch-Pagan test:

**Null hypothesis:** heteroscedasticity is not present ($H_0: \sigma^2_1 = \sigma^2_2 =... = \sigma^2_n$)

**Alternative hypothesis:** heteroscedasticity is present ($H_a:$ at least one $\sigma^2_i$ is different from the others)

We will set the alpha value to 0.05.

```{r}
bptest(model_deaths_final)
```

From the output of our test, we get a p-value of 0.00000000877 which is less than 0.05. This means we can reject the null hypothesis that there is homoscedasticity and conclude with a significance level of 0.05 that our model is homoscedastic. This means that there is a problem with the homoscedasticity assumption. 

### LINEARITY 

Testing for linearity using a residuals vs predicted $\hat{Y}$ plot:
```{r}
ggplot(model_deaths_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, our model does seem to be linear. ??????

### OUTLIERS

Testing for outliers using leverage:
```{r}
lev=hatvalues(model_deaths_final)
p = length(coef(model_deaths_final))
n = nrow(covid_agg)
outlier3p = lev[lev>(3*p/n)]
outlier3p
```

```{r}
plot(model_deaths_final,which=5)
```

From our output, there seems to be two data points, row 12, 44, 64, 76, 89, 99, 101, 107, and 110 with a leverage greater than $3p/n$. 

### INDEPENDANT ERRORS

Plotting residuals vs. fitted values to see if the error terms look independant from each other:
```{r}
ggplot(model_deaths_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```


Looking at the plot above, it seems that the residuals are evenly distributed, meaning that the residuals are independant from each other. ????????