---
title: "P603_Rodrigo"
author: "Rodrigo Rosales"
date: "2022-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide', echo=FALSE, include=FALSE}
# Import libraries
library(dplyr)
library(ggplot2)
library(mosaic)
library(tidyverse)
library(resampledata)
library(tidyr)
library(olsrr)
library(leaps)
library(mctest)
library(lmtest)
library(GGally)
library(agricolae)
library(MASS)
library(car)
options(scipen=999)
```

```{r}
# Importing COVDI-19 data set
covid_raw <- read_csv('data.csv', show_col_types = FALSE)
```

```{r}
# Getting countries with bottom 5% of new_cases_smoothed_per_million
tenth_percentile_cases <- quantile(covid_raw$new_cases_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_cases <- covid_raw[covid_raw$new_cases_smoothed_per_million < tenth_percentile_cases,]
bad_country_list_cases <- unique(bad_country_cases$location)

# Getting countries with bottom 5% of new_deaths_smoothed_per_million
tenth_percentile_deaths <- quantile(covid_raw$new_deaths_smoothed_per_million, probs = 0.05, na.rm = TRUE)
bad_country_deaths <- covid_raw[covid_raw$new_deaths_smoothed_per_million < tenth_percentile_deaths,]
bad_country_list_deaths <- unique(bad_country_deaths$location)

# Remove countries that appear in either above lists
bad_country_list <- append(bad_country_list_deaths, bad_country_list_cases)
good_countries <-   covid_raw[!covid_raw$location %in% bad_country_list, ]
```

```{r}
# change here to use final cleaned dataset
covid_data <- good_countries
covid_data$smokers <- (covid_data[['male_smokers']] + covid_data[['female_smokers']]) / 2

# dropping columns
covid = subset(covid_data, select = -c(iso_code, date, continent, handwashing_facilities, tests_units) )

# NaN to 0
covid[is.na(covid)] = 0

# Aggregation
covid_agg <- covid %>% group_by(location) %>% summarise(across(everything(), mean), .groups = 'drop') %>% as.data.frame()
```


# CASES MODEL
```{r}
model_cases_full = lm(new_cases ~ aged_65_older + smokers + cardiovasc_death_rate + diabetes_prevalence + extreme_poverty + gdp_per_capita + median_age + life_expectancy + population + stringency_index + human_development_index + reproduction_rate + hosp_patients + new_tests + population_density, data=covid_agg)

summary(model_cases_full)
```
```{r}
model_cases_stepwise = ols_step_both_p(model_cases_full, pent=0.1, prem=0.3, progress=TRUE)
```
```{r}
model_cases_revised = lm(new_cases ~ population + hosp_patients + life_expectancy, data=covid_agg)

summary(model_cases_revised)
```

```{r}
model_cases_interactions = lm(new_cases ~ (population + hosp_patients + life_expectancy)^2, data=covid_agg)
summary(model_cases_interactions)
```

```{r}
model_cases_interactions_2 = lm(new_cases ~ population + hosp_patients + life_expectancy + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_interactions_2)
```

```{r}
new_data <-data.frame(covid_agg$new_cases, covid_agg$population, covid_agg$hosp_patients, covid_agg$life_expectancy)

ggpairs(new_data, lower = list(continuous="smooth_loess", combo="facethist", discrete="facetbar", na="na"))
```

```{r}
model_cases_squared = lm(new_cases ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_squared)
```

```{r}
model_cases_final = lm(new_cases ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)

summary(model_cases_final)
```



# MODEL DEATHS
```{r}
model_deaths_full = lm(new_deaths ~ aged_65_older + smokers + cardiovasc_death_rate + diabetes_prevalence + extreme_poverty + gdp_per_capita + median_age + life_expectancy + population + stringency_index + human_development_index + reproduction_rate + hosp_patients + new_tests + population_density, data=covid_agg)

summary(model_deaths_full)
```

```{r}
model_deaths_stepwise = ols_step_both_p(model_deaths_full, pent=0.1, prem=0.3, progress=TRUE)
```

```{r}
model_deaths_revised = lm(new_deaths ~ population + hosp_patients + life_expectancy, data=covid_agg)

summary(model_deaths_revised)
```

```{r}
model_deaths_interactions = lm(new_deaths ~ (population + hosp_patients + life_expectancy)^2, data=covid_agg)
summary(model_deaths_interactions)
```


```{r}
model_deaths_interactions_2 = lm(new_deaths ~ population + hosp_patients + life_expectancy + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_interactions_2)
```

```{r}
new_data <-data.frame(covid_agg$new_deaths, covid_agg$population, covid_agg$hosp_patients, covid_agg$life_expectancy)

ggpairs(new_data, lower = list(continuous="smooth_loess", combo="facethist", discrete="facetbar", na="na"))
```

```{r}
model_deaths_squared = lm(new_deaths ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_squared)
```

```{r}
model_deaths_final = lm(new_deaths ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg)
summary(model_deaths_final)
```

# ASSUMPTIONS 

## CASES MODEL

### NORMALITY 

Testing for normality using the Shapiro-Wilk test:

**Null hypothesis:** the sample data are significantly normally distributed

**Alternative hypothesis:** the sample data are not significantly normally distributed

We will set the alpha value to 0.05.

```{r}
shapiro.test(residuals(model_cases_final))
```


Plotting a Q-Q plot:

```{r}
ggplot(model_cases_final, aes(sample=model_cases_final$residuals)) +
stat_qq() +
stat_qq_line()
```

From the output of our test, we get a p-value of 0.00000000000000022 which is less than 0.05. This means we can reject our null hypothesis that our sample data is significantly normally distributed and conclude with a significance level of 0.05 that our sample data is not normally distributed. This means that there is a problem with the normality assumption.

### HOMOSCEDASTICITY

Testing for heteroscedasticity using the Breusch-Pagan test:

**Null hypothesis:** heteroscedasticity is not present ($H_0: \sigma^2_1 = \sigma^2_2 =... = \sigma^2_n$)

**Alternative hypothesis:** heteroscedasticity is present ($H_a:$ at least one $\sigma^2_i$ is different from the others)

We will set the alpha value to 0.05.

```{r}
bptest(model_cases_final)
```

From the output of our test, we get a p-value of 0.0000001319 which is less than 0.05. This means we can reject the null hypothesis that there is homoscedasticity and conclude with a significance level of 0.05 that our model is not homoscedastic. This means that there is a problem with the homoscedasticity assumption. 

### LINEARITY 

Testing for linearity using a residuals vs predicted $\hat{Y}$ plot:
```{r}
ggplot(model_cases_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, our model does seem to be linear. ??????

### OUTLIERS

Testing for outliers using leverage:
```{r}
lev=hatvalues(model_cases_final)
p = length(coef(model_cases_final))
n = nrow(covid_agg)
outlier3p = lev[lev>(3*p/n)]
outlier3p
```

```{r}
plot(model_cases_final,which=5)
```

From our output, there seems to be two data points, row 12, 44, 64, 76, 89, 99, 101, 107, and 110 with a leverage greater than $3p/n$. 

Removing the rows that are influential outliers:
```{r}
covid_agg2 <- covid_agg[-c(12, 44, 64, 76, 89, 99, 101, 107, 110), ]
```


### INDEPENDANT ERRORS

Plotting residuals vs. fitted values to see if the error terms look independant from each other:
```{r}
ggplot(model_cases_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, it seems that the residuals are evenly distributed, meaning that the residuals are independent from each other. ??????

Since our model failed majority of the assumptions, we will try a Box-Cox transformation to see if it improves at all.

First, we should confirm that our response variable is always positive.

```{r}
covid_agg2[covid_agg2["new_cases"]<=0,]

# New model without outliers 
model_cases_final2 = lm(new_cases ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg2)
```

We get no rows where new cases is less than 0, so we can continue with the Box-Cox transformation.

```{r}
bc=boxcox(model_cases_final2,lambda=seq(-10,10))
bestlambda=bc$x[which(bc$y==max(bc$y))]
bestlambda
```

We get a lambda of 0.1010101. Conducting the Box-Cox transformation:

```{r}
bc_model_cases_final2 = lm((((new_cases^(-0.1010101))-1)/-0.1010101)~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy,data=covid_agg2)
```

## ASSUMPTIONS FOR BOX-COX TRANSFORMED MODEL

### NORMALITY 

Testing for normality using the Shapiro-Wilk test:

**Null hypothesis:** the sample data are significantly normally distributed

**Alternative hypothesis:** the sample data are not significantly normally distributed

We will set the alpha value to 0.05.

```{r}
shapiro.test(residuals(bc_model_cases_final2))
```


Plotting a Q-Q plot:

```{r}
ggplot(bc_model_cases_final2, aes(sample=bc_model_cases_final2$residuals)) +
stat_qq() +
stat_qq_line()
```

From the output of our test, we get a p-value of 0.005933 which is less than 0.05. This means we can reject our null hypothesis that our sample data is significantly normally distributed and conclude with a significance level of 0.05 that our sample data is not normally distributed. This means that there is a problem with the normality assumption.

### HOMOSCEDASTICITY

Testing for heteroscedasticity using the Breusch-Pagan test:

**Null hypothesis:** heteroscedasticity is not present ($H_0: \sigma^2_1 = \sigma^2_2 =... = \sigma^2_n$)

**Alternative hypothesis:** heteroscedasticity is present ($H_a:$ at least one $\sigma^2_i$ is different from the others)

We will set the alpha value to 0.05.

```{r}
bptest(bc_model_cases_final2)
```

From the output of our test, we get a p-value of 0.01903 which is less than 0.05. This means we can reject the null hypothesis that there is homoscedasticity and conclude with a significance level of 0.05 that our model is not homoscedastic. This means that there is a problem with the homoscedasticity assumption. 

### LINEARITY 

Testing for linearity using a residuals vs predicted $\hat{Y}$ plot:
```{r}
ggplot(bc_model_cases_final2, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, our model does seem to be linear. ??????

### INDEPENDANT ERRORS

Plotting residuals vs. fitted values to see if the error terms look independant from each other:
```{r}
ggplot(bc_model_cases_final2, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```


## DEATHS MODEL

### NORMALITY 

Testing for normality using the Shapiro-Wilk test:

**Null hypothesis:** the sample data are significantly normally distributed

**Alternative hypothesis:** the sample data are not significantly normally distributed

We will set the alpha value to 0.05.

```{r}
shapiro.test(residuals(model_deaths_final))
```

Plotting a Q-Q plot:

```{r}
ggplot(model_deaths_final, aes(sample=model_deaths_final$residuals)) +
stat_qq() +
stat_qq_line()
```

From the output of our test, we get a p-value of 0.00000000000000022 which is less than 0.05. This means we can reject our null hypothesis that our sample data is significantly normally distributed and conclude with a significance level of 0.05 that our sample data is not normally distributed. This means that there is a problem with the normality assumption.

### HOMOSCEDASTICITY 

Testing for heteroscedasticity using the Breusch-Pagan test:

**Null hypothesis:** heteroscedasticity is not present ($H_0: \sigma^2_1 = \sigma^2_2 =... = \sigma^2_n$)

**Alternative hypothesis:** heteroscedasticity is present ($H_a:$ at least one $\sigma^2_i$ is different from the others)

We will set the alpha value to 0.05.

```{r}
bptest(model_deaths_final)
```

From the output of our test, we get a p-value of 0.00000000877 which is less than 0.05. This means we can reject the null hypothesis that there is homoscedasticity and conclude with a significance level of 0.05 that our model is homoscedastic. This means that there is a problem with the homoscedasticity assumption. 

### LINEARITY 

Testing for linearity using a residuals vs predicted $\hat{Y}$ plot:
```{r}
ggplot(model_deaths_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, our model does seem to be linear. ??????

### OUTLIERS

Testing for outliers using leverage:
```{r}
lev=hatvalues(model_deaths_final)
p = length(coef(model_deaths_final))
n = nrow(covid_agg)
outlier3p = lev[lev>(3*p/n)]
outlier3p
```

```{r}
plot(model_deaths_final,which=5)
```

From our output, there seems to be two data points, row 12, 44, 64, 76, 89, 99, 101, 107, and 110 with a leverage greater than $3p/n$. 

Since the rows that are influential outliers are the same for our cases model, we will use the same dataset with removed outliers (covid_agg2)

### INDEPENDANT ERRORS

Plotting residuals vs. fitted values to see if the error terms look independant from each other:
```{r}
ggplot(model_deaths_final, aes(x=.fitted, y=.resid)) +
geom_point() + geom_smooth()+
geom_hline(yintercept = 0)
```

Looking at the plot above, it seems that the residuals are evenly distributed, meaning that the residuals are independant from each other. ????????

```{r}
covid_agg2[covid_agg2["new_cases"]<0,]
model_deaths_final2 = lm(new_deaths ~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy, data=covid_agg2)
```

We get no rows where new cases is less than 0, so we can continue with the Box-Cox transformation.

```{r}
bc=boxcox(model_cases_final2,lambda=seq(-10,10))
bestlambda=bc$x[which(bc$y==max(bc$y))]
bestlambda
```

We get a lambda of -0.1010101. Conducting the Box-Cox transformation:

```{r}
# Getting error when running lm about having Inf in the data frame. 
bcmodel_deaths_final2 = lm((((new_deaths^(-0.1010101))-1)/-0.1010101)~ population + hosp_patients + life_expectancy + I(life_expectancy^2) + population:life_expectancy + hosp_patients:life_expectancy,data=covid_agg2)

# There are no NAs in new_deaths or the Box-Cox transformation
all(is.na(covid_agg2$new_deaths))
all(is.na((((covid_agg2$new_deaths^(-0.1010101))-1)/-0.1010101)))

# There are no Inf in new_deaths or the Box-Cox transformation
all(is.infinite(covid_agg2$new_deaths))
all(is.infinite((((covid_agg2$new_deaths^(-0.1010101))-1)/-0.1010101)))
```

When trying to conduct a Box-Cox transformation on our new_deaths model, we encounter an error. After some testing we are not sure why this error is occuring, so we will leave the model without a Box-Cox transformation.